Appendix

### 2025-02-18 manuscript_rmarkdown/manuscript_appendix.Rmd
Revise appendix F.

An alternative to location- and scale-invariant transformations is normalising feature distributions prior to conventional transformations.
Table \ref{tab:normalisation-before-transformation-appendix} shows residual errors, after transformation to normality, of the five features from real-world datasets presented previously in Appendix D and the main manuscript.
In these examples location- and scale-invariant transformations have similar or lower residual errors compared to errors resulting from normalisation prior to transformation.

@@

data <- .assess_standardisation_before_transformation()
data <- data[transformation %in% c("none", "conventional", "conventional (z-score norm.)", "conventional (robust scaling)", "invariant"), mget(c("residuals", "transformation", "dataset"))]
data[, "residuals" := round(residuals, digits = 1L)]
data.table::dcast(data, dataset ~ transformation, value.var = "residuals")
```
```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. conv.: conventional; norm.: normalisation; rob.: robust}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & none & conventional & conv. (z-score norm.) & conv. (rob. scaling) & invariant \\

\midrule
age                     &  16.5 &  11.5 &  11.5 &  11.3 &   8.8 \\
penguin body mass       &  48.0 &  32.2 &  33.3 &  32.2 &  26.8 \\
latitude                & 328.1 & 319.0 & 326.2 & 324.5 & 326.4 \\
fuel efficiency         &  54.5 &  55.3 &  49.0 &  53.3 &  44.0 \\
arterial wall thickness & 110.1 &  30.0 &  19.3 &  31.8 &  12.2 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

>>>>

An alternative to location- and scale-invariant transformations is normalising feature distributions prior to conventional transformations.
We assessed two normalisation methods were used:
1. z-standardisation, i.e. $x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma$, with $\mu$ and $\sigma$ the mean and standard deviation of feature $\mathbf{X}$; 
and 2. robust scaling, i.e. $x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / \text{IQR}\left(\mathbf{X}\right)$, with $\text{IQR}$ representing the interquartile range of feature $\mathbf{X}$.

Table \ref{tab:normalisation-before-transformation-appendix} shows residual errors after transformation to normality of the five features from real-world datasets presented previously in Appendix D and the main manuscript.
Table \ref{tab:normalisation-before-transformation-lambda-appendix} shows the corresponding values for transformation parameter $\lambda$.
Normalisation prior to transformation prevents unexpectedly high $\lambda$ values in the presence of large location shifts, as can be observed for the latitude feature.
Normalisation prior to transformation, however, does not consistently improve residual errors after transformation. 
Interestingly, use of normalisation prior to transformation with robust power transformation using the method by Raymaekers and Rousseeuw [@Raymaekers2024-zf] led to higher residual errors for arterial wall thickness, fuel efficiency and latitude features.
This is similar to observations by @Raymaekers2024-zf, who observed that their method can emphasize values of outliers after transformation.
Location- and scale-invariant transformations have similar or lower residual errors compared to errors resulting from normalisation prior to transformation.

@@

data <- .assess_standardisation_before_transformation()
data <- data[, mget(c("residuals", "lambda", "transformation", "dataset"))]
data[, "residuals" := round(residuals, digits = 1L)]
data[, "lambda" := round(lambda, digits = 1L)]
data.table::dcast(data, transformation ~ dataset, value.var = "residuals")
data.table::dcast(data, transformation ~ dataset, value.var = "lambda")
```

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & age & AWT & FE & latitude & PBM \\

\midrule
none & 16.5 & 110.1	& 54.5 & 328.4 & 48.0 \\
conventional & 11.5 & 30.0 & 55.3 & 319.0 & 32.2 \\
conventional (z-score norm.) & 11.5 & 19.3 & 49.0 & 326.2 & 33.3 \\
conventional (robust scaling) & 11.3 & 31.8 & 53.3 & 324.5 & 32.2 \\
Raymaekers-Rousseeuw & 11.5 & 136.7 & 47.7 & 315.1 & 32.2 \\
Raymaekers-Rousseeuw (z-score norm.) & 13.2 & 214.5 & 57.4 & 326.2 & 33.3 \\
Raymaekers-Rousseeuw (robust scaling) & 11.7 & 281.3 & 57.0 & 324.5 & 32.0 \\
invariant & 8.8 & 12.2 & 44.0 & 326.4 & 26.8 \\
robust invariant & 10.6 & 27.1 & 49.5 & 310.6 & 23.1 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & age & AWT & FE & latitude & PBM \\

\midrule
conventional & 2.0 & -0.7 & -0.1 & 62.1 & -0.5 \\
conventional (z-score norm.) & 1.2 & -1.7 & -0.1 & 1.3 & 0.6 \\
conventional (robust scaling) & 1.3 & 0.0 & 0.2 & 1.4 & 0.4 \\
Raymaekers-Rousseeuw & 2.0 & 1.1 & 0.8 & 95.4 & -0.5 \\
Raymaekers-Rousseeuw (z-score norm.) & 1.3 & 1.4 & 1.0 & 1.3 & 0.6 \\
Raymaekers-Rousseeuw (robust scaling) & 1.4 & 1.3 & 1.0 & 1.4 & 0.4 \\
invariant & 1.3 & 0.2 & -1.3 & 1.5 & 0.5 \\
robust invariant & 1.4 & -0.6 & 0.9 & 1.4 & 0.4 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```
